---
title: "Reading FIT files"
author: "Mike L. Smith"
date: "`r Sys.Date()`"
vignette: |
  %\VignetteIndexEntry{Reading FIT files}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, out.width="100%")
```

# Introduction

**FITfileR** is an R package to read FIT files produced by fitness tracking devices like Garmin Edge cycle computers or sports watches. The intention for **FITfileR** is to use native R code to read the files directly, with no reliance on the FIT SDK or other FIT parsing tools. As such it should be platform independent, and not require any additional software outside of a working version of R.

## Installing and loading the library

Currently **FITfileR** is only available on Github, and can be installed using the **[remotes](https://cran.r-project.org/package=remotes) package.**

```{r, eval = FALSE}
if(!requireNamespace("remotes")) {
    install.packages("remotes")
}
remotes::install_github("grimbough/FITfileR")
```

Once the package is installed, you then need to load the library in your R session before it can be used.

```{r load_lib}
library(FITfileR)
```

## Example fit files

**FITfileR** is distributed with several example FIT files to test its functionality.  These files can be found in the `extdata\Activities` folder of the package, and you can see the names of all the example files with:

```{r}
list.files(system.file("extdata", "Activities", package = "FITfileR"))
```

The names of these files indicate the manufacture and model name of the device it was recorded on, as well as the activity type.  There are many other differences (e.g. connected sensors, recording frequency, software versions) that are not encapsulated in the file names.

# Reading files

To demonstrate reading a FIT file we're going to use the `garmin-fenix6-swim.fit` file distributed with **FITfileR**.

```{r data_files, cache = TRUE}
library(FITfileR)
fenix6_file <- system.file("extdata", "Activities", "garmin-fenix6-swim.fit", 
                            package = "FITfileR")
```

We read files using the function `readFitFile()`.

```{r reading_files}
fenix6 <- readFitFile(fenix6_file)
```

The resulting object is an object of type `FitFile` containing all the data stored in the original FIT file.  Typing the name of the object will print some details about the file e.g. the time it was created, the manufacturer and name of the device it was recorded on, and the number of data 'messages' held within the file.  *Exactly what is shown here will depend on the information available in the file, so it may look slightly different for you*.

```{r show}
fenix6
```

# Working with the data

If we want to do more than just print a summary of the FIT file to screen we need to use some accessor function to extract the data from our FitFile object.  There are several ways to achieve this depending on the datatype you're interested in.

## Records - GPS, speed, altitude, etc

The data most often wanted from a fit file are the values such as location, speed, altitude, etc recorded during an activity.  Such data are classed as *records* in the FIT specification and can be retrieved from the FitFile object using using `records()`. 

```{r print_record}
fenix6_records <- records(fenix6)
fenix6_records
```

In this example we actually get a list with three `tibbles`.  This is because in this particular file there are three distinct definitions of what a *record* contains.  This normally happens if data recording begins before a sensor (e.g. a heart rate monitor) has been attached to a device, or GPS position has been acquired, although sometimes the reason can be more opaque.  In this example it seems clear that we can use the second entry, which contains the vast majority of the data.  *Note: sometimes the bulk of your data may be spread across multiple `tibbles` in the list rather than a single entry.  See the "Plotting a route" section below for an example of how to handle this.*

## Extracting common data types

In addition to the `records()` function, **FITfileR** provides an number of other methods for accessing commonly found message types.  Currently, these include:

  - `laps()`
  - `events()`
  - `file_id()`
  - `hrv()`
  - `monitoring()`

## Accessing any data type

The FIT specification allows for `r nrow(FITfileR:::fit_data_types$mesg_num)` distinct message types, and **FITfileR** does not include specific accessor functions for each of these.  To view a complete list of the message types stored within a file you can use the function `listMessageTypes()`.

```{r listMessageTypes}
listMessageTypes(fenix6)
```

We can see there are `r length(listMessageTypes(fenix6))` different message types in the file above.  If a specific accessor method doesn't exist for the message type you're interested in, you can use the function `getMessagesByType()` and provide the message type name to the `message_type` argument.  The code below will extract all "zones_target" messages from our file.  

```{r devicesettings}
getMessagesByType(fenix6, message_type = "zones_target")
```

In this case *zones_target* is a single message that reports power and heart rate thresholds that were set on the device.  One could imagine using these in conjunction with the *records* to measure how well the athlete performed relative to the pre-set threshold for this particular activity.

# Example use cases

## Plotting a route

```{r, read-edge530, cache = TRUE}
edge530_file <- system.file("extdata", "Activities", "garmin-edge530-ride.fit", 
                            package = "FITfileR")
edge530 <- readFitFile(edge530_file)
```

To plot locations we extract the longitude and latitude from our FIT records.  These data are found in *record* messages, and we use `records()` to extract them.  As before, we are returned a list of tibbles.  However, unlike the previous example there is no entry that clearly holds almost all the data; there are two different definitions for *record* messages with over one thousand data points.  

```{r process-records}
edge530_records <- records(edge530)

## report the number of rows for each set of record messages
vapply(edge530_records, FUN = nrow, FUN.VALUE = integer(1))
```

We probably don't want to discard either of these, as even the smaller one represents over 20 minutes of data recording.  We can use **dplyr** to try and merge all the messages together into a single `tibble` regardless of their definition.  Any entries that are missing in certain messages will be filled with `NA`.  *Note: this approach of binding rows does not always work, as sometimes the data types within a column may change between messages, but it is more often successful.*

```{r merge-records, message=FALSE}
library(dplyr)

edge530_allrecords <- records(edge530) %>% 
  bind_rows() %>% 
  arrange(timestamp) 

edge530_allrecords
```

We can then use `dplyr::select()` to extract the latitude and longitude columns from our `tibble`, so we can pass them easily to a plotting function.

```{r, extract-coords}
coords <- edge530_allrecords %>% 
  select(position_long, position_lat)
```

We can now use the **leaflet** package to create an interactive map, with our route overlayed on top.

```{r mapping, message=FALSE}
library(leaflet)

m <- coords %>% 
  as.matrix() %>%
  leaflet(  ) %>%
  addTiles() %>%
  addPolylines( )
    
m
```

## Comparing heart rate measurments between devices

The package comes with two example fit files, recorded during a Zwift activity in 2021.   They are of the same ride and record the same rider, but some of the data recording was carried out on two different devices so they could be compared. Specifically, a Garmin heart rate strap and and Elite Direto trainer were linked to Zwift for recording heart rate and power, alongside a Garmin Fenix 6 wrist based heart rate monitor and Assioma Duo pedals measuring the same properties.  

Here we compare the heart rates recorded with two devices, to see how consistent they are with each other.

```{r, heart-rate-packages}
library(dplyr)
library(tidyr)
library(ggplot2)
```

First we need to locate the two files and read them into R:


```{r read-both, cache = TRUE}
garmin_file <- system.file("extdata", "Activities", "garmin-fenix6-turbo.fit", 
                            package = "FITfileR")
zwift_file <- system.file("extdata", "Activities", "zwift-turbo.fit", 
                            package = "FITfileR")

garmin <- readFitFile(garmin_file)
zwift <- readFitFile(zwift_file)
```

We then use `records()` to extract the appropriate messages from the two files.  Since there's a lot of data in addition to the heart rate readings we're interested in, we use functions from **dplyr** and **tidyr** to combine the heart rate data into a single data frame, and make it into a 'long' format suitable for plotting with **ggplot2**.

```{r read-hr, eval = TRUE}
garmin_records <- records(garmin) %>%
    bind_rows() %>% 
    arrange(timestamp) 
zwift_records <- records(zwift)

hr_table <- inner_join(garmin_records, zwift_records, by = "timestamp") %>%
  select(timestamp, Garmin = heart_rate.x, Zwift=heart_rate.y) %>%
  tidyr::pivot_longer(cols = Garmin:Zwift, names_to = "device", values_to = "heart_rate")
```

```{r plot-hr, fig.width=8, fig.height=3}
ggplot(hr_table) + 
    geom_line(aes(x = timestamp, y = heart_rate, col = device)) +
    theme_bw()
```

The two heart rate monitors correspond well over the duration of the workout.  There seems to be a 1-2 second lag in the wrist-based monitor (Garmin), but the high and low values track really well.  Based on this I'd have no issue relying on the Fenix 6 heart-rate monitor for my training.

We can also consider the difference between the two measurements both at each common time point and also as a rolling mean over a 60 second window.

```{r, message = FALSE, warning = FALSE, fig.width=8, fig.height=3}
library(zoo)

hr_differences <- hr_table %>% 
  pivot_wider(id_cols = timestamp, names_from = device, values_from = heart_rate) %>% 
  mutate(hr_diff = Garmin - Zwift,
         Garmin_hr_zone = cut(Garmin, breaks = seq(0,200,20))) %>%
  mutate(hr_60 = zoo::rollmean(hr_diff, k = 60, fill = NA))

ggplot(hr_differences, aes(x = timestamp, y = hr_diff)) + 
  geom_point(aes(col = Garmin_hr_zone)) +
  geom_line(aes(y = hr_60), col = "grey40", lwd = 1.6, alpha = 0.7) + 
  geom_abline(intercept = 0, slope = 0) + 
  theme_bw() +
  ylab("Heart Rate Difference\n+ve Chest Strap Greater / -ve Wrist Watch Greater") +
  scale_colour_brewer(palette = "PRGn")
```

For the most part the readings are quite similar between the two devices.  Just looking at the plot, it seems there are more points above the zero line, indicating that the chest strap records slightly higher values on average, but the difference is generally within 5 beats-per-minute.  Based on the colouring it looks like the largest differences occur when the heart-rate is high.

# More details

In this section we describe a few more of the implementation details of **FITfileR** and some of the choices made in presenting the data in an R session.

## Data types and units

Much of the data contained in FIT files is not stored in the formats that you might instinctively expect or that see on the display of the device the file was recorded on.  For example, the elapsed time of an activity isn't stored in seconds, but rather milliseconds, and this value needs to be scaled to get the time in seconds.  Other data types require more complex processing.  The latitude and longitude positional information, which is not stored as decimal degrees (e.g. `42.87342`) but rather as a signed 32-bit integer (e.g. `511512370`) representing "semicircles", needs to be converted by multiplying by the scaling factor $180 / 2^{31}$.  Similarly, text information such as the activity type or device manufacturer isn't stored directly as the string "run" or "Garmin" but as an integer that maps to an entry in a table of sports or manufacturers respectively.

More details of the data types can be found in the *Profile.xlsx* file that is provided as part of the [FIT SDK](https://developer.garmin.com/fit/download/).  **FITfileR** has its own internal representation of this file, and will try to convert many data types automatically.  In most cases it is possible to find the units for the values **FITfileR** is displaying via the `units` attribute.  This will either be printed to screen alongside the contents of a data column, or you can extract the `units` attribute directly.

```{r, eval = TRUE, echo = TRUE}
garmin_session <- getMessagesByType(garmin, "session")
## show the latitude of the start position
garmin_session$start_position_lat
## extract the units for the total ascent value
attr(garmin_session$total_ascent, "units")
```

This automatic conversion works for many of the more common fields based around timestamps, GPS positions, distances, speeds, and probably many more.  However the FIT specification is large and it is likely there are common data types that I have not encountered in one of my own FIT files.  If you come across a field that is entirely `NA` then it is likely that your file includes a data type that **FITfileR** does not currently support.  Please open an issue at [GitHub](https://github.com/grimbough/FITfileR/issues) and I will try to add the required functionality.

# Session Info

```{r}
sessionInfo()
```