---
title: "Reading .fit files"
author: "Mike Smith"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load_lib}
library(fitFileR)
```

# Reading files

## Example fit files

The package comes with two example fit files, recorded during a ride in early 2017.  They are of the same ride and record the same rider, but the data logging was carried out on two different devices: a Garmin Edge 500 and a TomTom Runner 3.  The files can be found in the `extdata` folder of the package, and identified on your system using the following code. 

```{r data_files}
garmin_file <- system.file("extdata/Garmin.fit", package = "fitFileR")
tomtom_file <- system.file("extdata/TomTom.fit", package = "fitFileR")
```

## Reading

We read files using the function `readFitFile()`.

```{r reading_files}
garmin <- readFitFile(garmin_file)
tomtom <- readFitFile(tomtom_file)
```

The resulting object is a `list` of `tibbles` containing all the data stored in the fit file.

```{r exploring_list}
names(garmin)
```

# Working with the data

The data most often wanted from a fit file are the values such as location, speed, altitude, etc recorded during an activity.  These can be found in the `record` entry in the list.  Here's the first few rows from our TomTom file:

```{r print_record}
garmin$record
```

## Plotting a route

Before we can plot locations need to process the values held in the fit file into longitude and latitude.  First, we remove any entries recorded when satelite reception was lost; these have a value of `NA` in this file, but appear as `2^31-1` in the data from the TomTom so we remove both.  We then convert to longitude and latitude coordinates.

```{r process_coords}

coords <- cbind(garmin$record$position_long, garmin$record$position_lat)
coords <- coords[ apply(coords, 1, function(x) { !any(is.na(x) || x == 2^31-1) }), ]
coords <- coords * (180 / (2^31))
```

We can now use the **leaflet** package to create an interactive map, with our route overlayed on top.

```{r mapping}
library(leaflet)

m <- leaflet( coords ) %>%
     addTiles() %>%
     addPolylines( )
    
m
```

## Comparing heart rates

```{r plot_hr}
ggplot() + 
    geom_line(data = garmin$record, aes(x = timestamp, y = heart_rate), col = "blue") + 
    geom_line(data = tomtom$record, aes(x = timestamp, y = heart_rate), col = "red")

ggplot() + 
    geom_line(data = garmin$record, aes(x = timestamp, y = heart_rate), col = "blue") + 
    geom_line(data = dplyr::filter(tomtom$record, heart_rate != 255), 
              aes(x = timestamp, y = heart_rate), col = "red")
```